Application and Relational Database Co-Refactoring

### 2013-03-04 doc/paper.tex
text improvements

\title{Refactoring of Application Using Relational Database}

@@

\begin{abstract}
The change of application source code caused by refactoring affected not only the application, but it affect the stored data too if the persistent objects change. The change is usually processed in two steps: refactoring and data migration. We provide a solution capable to migrate database according to a refactoring in the application code. The feasibility of change and its data-secure processing is addressed too.

@@

\subsection{Solution Proposal}
We believe there exists a better solution which automatizes the process of database evolution according to the code refactoring. The solution is illustrated in the Fig.~\ref{fig:evolution}. It is based on a change in the evolution process which assumes that the ORM does not change during the evolution:
\begin{enumerate}
	\item The evolution of the whole software is defined independently on the application or the database.
	\item The evolution is interpreted for the application and the database.
	\item The evolution is executed.
\end{enumerate}
The process can decrease the human mistakes in the evolution process, because there is only one source for the evolution and the evolution is automatically interpreted for the application (as a refactoring) and database (as schema and data migration). The existence of the set of all possible software evolutions $E$ is based on a evolutionary transformations specific for application and database. Each transformation contains conditions of transformation feasibility, thus the feasibility of the evolution can be verified.

@@

\paragraph{Example 2} The same situation from Example 1 can be solved more effectively if the \emph{inline} evolution is defined in $E$. All information needed for software evolution are provided as the input of the \emph{inline} transformation which is then automatically interpreted as inline of application classes and merge of tables on database level. The mapping is provided as the input of \emph{inline} too, thus the script for data migration can be generated. The structural feasibility is verified during the interpretation of the \emph{inline} and the migration script contains conditions which verifies the feasibility on level of data.

@@

\section{Model of Software Evolution}
\label{sec:models}
The model-driven development (MDD) is a good approach to data evolution of various software components \cite{Moravec:PracticalApproach}. The software is represented by a set of models - application model and database model concretely. The evolution and ORM is represented by a set of model-to-model transformations and the interpretation of transformations from $E$ is a model-to-model transformation again.

Three sets of transformations has to be defined to provide the capabilities described in Example 2: the set $E$ of all possible software transformations:
\begin{equation}
E = \{t | t: Software \rightarrow Software\}\:,
\end{equation}
the set of application refactorings:
\begin{equation}
A = \{t | t: Application \rightarrow Application\}
\end{equation}
and the set of database evolutionary transformations
\begin{equation}
 D = \{t | t: Database \rightarrow Database\}\:.
\end{equation}
The evolution of the software is defined as the interpretation of the transformation for each component of the software:
\begin{align}
& t(s) = \begin{cases}
software(\Psi(s, t), \Phi(s,t)) \; \mathbf{if} \; s \neq \perp \\\\
 	\perp
\end{cases}
s \in Software, t \in E \nonumber
\end{align}
Where $\Psi : E \rightarrow A $ interprets the software evolution cases to the code refactoring and $\Phi : E \rightarrow D $ interprets to the evolutionary transformation of database. The $\perp$ symbol denotes an inconsistent state of a software or its component. The definition of evolution allows to concatenate evolutionary transformations easily.

\subsection{Software Model}
The software consists of three important components (as seen in Fig. \ref{fig:evolution}: application (its persistent layer concretely), database (consisting of database schema and stored data) and an ORM, therefore we define a software as a triple consisting of an application and a database which are connected together by an ORM.
The software has to be in consistent state so its users could have benefit from its usage. The software is in consistent state if application and database are consistent and the database structure corresponds to the application structure according to the ORM, therefore the software is defined as:
\begin{align}
& software(a, d, \rho) = \begin{cases}
 software(a, d, \rho) \; \textbf{if} \; a \neq \perp \; \wedge \; b \neq \perp \\ \wedge \; \rho(a) = schema(d) 
 \\\\
 \perp
 \end{cases}\\
& a \in Application, d \in Database, \rho \in ORM \nonumber
\end{align}
The evolution of a software is a transformation from one consistent state to another one. These states are called generations of the software and the function which changed the state of a software are called transformations.

@@

\subsection{Application Model}
Application is defined as a sequence of classes and it creates the context for all structures used in the software persistent layer:
\begin{align}
& \mathbf{Application} = (Class*) \\
& 	\mathbf{Class} = (label, Property*, Association*, Inheritance) \\
& \mathbf{Property} = (label, AppType, DefaultValue, \nonumber \\ & \;\;\;\;  Cardinality, Mandatory) \\
&	\mathbf{Association} = (label, classRef, startCardinality, endCardinality)  \\
& \mathbf{Inheritance} = (class, InheritanceType) \\¨
& \mathbf{InheritanceType} = SINGLETABLE \; | \; TABLE-PER-CLASS \\
& \mathbf{AppType} = APPSTRING
\end{align}

\paragraph{Class} Class represents a basic structural unit in the application model. It has a unique name, one or more properties and it can be associated to other classes in the application. The properties of a class can be of two kinds - values or collections - according to theirs cardinality.

\paragraph{Property} Property represents a feature of a class which is represented as a primitive type. The property can be mandatory, can have a default value and according to its cardinality it can represent a single value or a collection of values.

\paragraph {Association} Association represents a connection between two classes. It has a unique name and the reference is represented by the label of referenced class. The class which owns the association is consider to be the starting class of an association, referenced class is consider to be the ending class of an association. The cardinalities defines the multiplicity of both association ends.

@@

 There is not allowed multiple ancestors and of course a class can not be its own ancestor cyclically. We also assume there is only one type of inheritance type in the class hierarchy.

\subsection{Application Manipulation}
\label{sec:app-evolution}
The application model defines only the structure of an application thus there are defined transformations for its creating, altering and removing parts of the model (the set $A$). The conditions for successful transformations are simple in case of creating or altering:
\begin{enumerate}
	\item The name collisions has to be prevented when creating an class in the model or altering its name.
	\item All references have to be updated when renaming a class.
	\item The existence of referenced class has to be verified when creating or altering an association.  
\end{enumerate}
The only issue during deleting classes etc. from an application is the class can be removed only in case it is not associated by other classes. The transformations from $A$ are in table \ref{tab:app-evolution}. The list of altering operations is not complete as there are missing transformation for changing the obligation or cardinality of properties and associations. It is because these transformations are not so important for complex refactorings. 
\begin{table}
\caption{The transformations for application evolution.}
	\label{tab:app-evolution}
\centering
	\begin{tabular}{ll}
	\hline
	Type of transformation & Defined transformations \\
	\hline
	\\[-2ex] 
	Application Creation
	& newProperty(className, property, application) \\
	& newAssociation(className, association, mapping, \\ &  \hspace{0.5in}application) \\
	&  newClass(class, application) \\
	Application Alternation
		& renameProperty(className, oldName, newName, \\ &  \hspace{0.5in} application) \\
		& renameAssociation(className, oldName, newName, \\ &  \hspace{0.5in}application) \\
		& renameClass(oldName, newName, application)\\
		& changeReference(referncedClass, newReferencedClass) \\
		Application Disassembly 
		& removeProperty(className, propertyName,\\ &  \hspace{0.5in} application) \\
		& removeAssociation(className, associationName, \\ &  \hspace{0.5in}application) \\
		& removeClass(className, application)\\
%	Copy Structure
%		& copyProperty(sourceClass, property, targetClass,\\ &  \hspace{0.5in} newName, application)  \\
%		& copyClass(class, newName, application) \\
	Inheritance manipulation   & setParent(class, parentClass) \\
		& removeParent(childClass) \\
		& pushDown(class, property)\\
		& pullUp(class, property)\\
	\hline
	\end{tabular}
\end{table}

\subsection{Database Model}
The relational database consists of database schema which defines the structure of the database and data, which in the software represents stored instances. The database is defined as:
\begin{align}
&	\mathbf{Database} = ( Table*, Row* )\\
&	\mathbf{Table} = (label, primaryKey, Column*, ForeignKey*)\\
&	\mathbf{Column} = (label, DbType, DefaultValue, Constraint*) \\
&	\mathbf{ForeignKey} = (Label, Label, Constraint*) \\
&	\mathbf{PrimaryKey} =  ( Label, Sequence ) \\
&	\mathbf{DbType} = DBSTRING \; | \; DBINT\\
&	\mathbf{Constraint} = NOTNULL \; | \; UNIQUE 
\end{align}

\paragraph{Table} Table represents a basic concept of database schema. It has a unique name, one or more columns and it can be related to other tables in the schema by foreign keys. Rows in the table represents stored data.

@@

\paragraph{Primary key} Primary key is unambiguous identifier of a record in a table. The primary key is always provided (automatically generated) by the database as a non-zero natural number. Primary key is always defined with constraints\\ $NOTNULL$ and $UNIQUE$. The sequence generates values of primary key. The new value of a key is obtained by calling the function $next(s)$. The sequence which provides numbers incremented by one is called $s_1$ if the increment is 1, $s_{10}$ if the increment is 10 etc.

@@

\paragraph{Data} A database consists not only from schema but also from data which are represented as rows in a table. A table row in our model is a tuple consisting of value pairs, which represents concrete value of concrete column or key. Each row contains a name of a table it belongs to.
\begin{align}
&	\mathbf{Row} = (Label, Pair*) \\
&	\mathbf{Pair} = (Label, Value) 
\end{align}

\subsection{Database Manipulation}
\label{sec:db-evolution}
The database consists of two parts - from database schema which defines the structure and from stored data - hence the transformations from set $D$ have to consider both parts. The transformation for manipulation of structure has similar conditions to the evolution of application. The transformations for data manipulation are inspired by the SQL language. The basic transformations are in Table \ref{tab:db-basic-evolution}. The set of transformations for database manipulation is limited in contrast with the SQL language as we mentioned only transformation necessary for data evolution on database level.

\begin{table}
\caption{The transformations for database evolution. The transformations can be mapped to SQL intuitevly.}
	\label{tab:db-basic-evolution}
\centering
\setlength\tabcolsep{0.5em}
	\begin{tabular}{ll}
	\hline
	Type of transformation & Defined transformations \\
	\hline
	\\[-2ex] Database Creation
	& addColumn(tableName, column, database) \\
	& addForeignKey(tableName, key, database) \\
	& addTable(table, database)\\
	Database Alternation
	& alterColumnName(tableName, oldName, \\ & \hspace{0.5in}newName, dataabse) \\
	& alterForeignKeyName(tableName, oldName, \\ & \hspace{0.5in}newName, dataabse) \\
	& alterTableName(oldName, newName, dataabse) \\
	Database Disasembling
	& dropColumn(tableName, columnName, database) \\
	& dropForeignKey(tableName, keyName, database) \\
	& dropTable(tableName, database) \\
	Data Manipulation
	& select(tableName, rowId, database) \\
	& select(tableName, database) \\
	& insert(row, database) \\
	& insert(rowId, value, database) \\
	& delete(tableName, rowId, database) \\
	& delete(tableName, rowId, columnName, database) \\
	Copy Structure
	& copyColumnStructure(sourceTableName,\\ &  \hspace{0.5in} targetTableName,  sourceColumnName, \\ &  \hspace{0.5in} newName, database) \\
	& copyTableStructure(sourceTable, newTableName, \\ &  \hspace{0.5in} database) \\
	Copy Structure and Values
	& copyColumn(sourceTable, targetTable, \\
	& \hspace{0.5in} sourceColumnName, targetColumnName, \\ 
	& \hspace{0.5in} mapping, database) \\
	& copyTable(sourceTable, newName, database) \\
	Data-secure Database Elements &
	dropEmptyColumn(tableName, columnName, \\
	Removal & \hspace{0.25in} database) \\
	& dropEmptyForeignKey(tableName, keyName, \\ & \hspace{0.25in} database) \\
	& dropEmptyTable(tableName, database) \\
	\hline
	\end{tabular}
\end{table}

\paragraph{Copying Database Elements} The transformations for copying the structure an values of a column or table serve more as helpers for advanced evolution cases, where they are discussed in detail.

@@

\paragraph{Data-secure Database Elements Removal} The transformation for remove elements from the database can have fatal impact on the data preservation, therefore the set of transformations is extended by a data secure transformations for removing database elements. These transformations are not part of the SQL standard, although they can be implemented as database functions. These transformations creates a secure way to remove elements from the database as they drops empty structural elements only.

\subsection{Object Relational Mapping}
The ORM is the only fixed point in the software model we use. The mapping is similar to the Hibernate mapping \cite{Hibernate} , thus a lot of developers should be familiar with it. The main ideas are: 
\begin{itemize}
	\item classes are mapped to tables,
	\item single properties are mapped to column or if the property is collection then the property is mapped to a table,
	\item associations are mapped to foreign keys or to tables if the association represents many-to-many  relationship,
	\item the primary keys are created automatically for each table,
	\item names used on application are mapped into the database schema (e.g. because possible name collision of application classes and names of database schema elements) 
\end{itemize}
%The ORM mapping is a function $ORM : Application \rightarrow Database $.

@@

\subsection{Basic Evolutionary Transformations}
\label{sec:sw-basic-evolution}
The evolution of the whole software is based on the atomic transformations specific for each software part, which are defined in Sec. \ref{sec:app-evolution} and \ref{sec:db-evolution}. This section introduces how these primitives can be used to manipulate the software elements.

The basic evolutionary transformations of the software are based on basic evolution of an application. These transformations has to respect the ORM, because properties and associations can be mapped as a column (foreign key respectively) as shown on the example of creating a new property: 
\begin{align}
& \Phi(newProperty(className, property, application)) = \nonumber \\
& \;\;\; =  \begin{cases}
  addColumn(ORM(className), ORM(property), database) \\\mathbf{if} \; cardinality(property \leq 1)  \\\\
  addTable(ORM(property), database) \\
  \mathbf{if} \; cardinality(property > 1)  
   \end{cases}
\end{align}

Mapping between basic evolutionary transformations is provided in the shortened way in table \ref{tab:sw-basic-evolution}, there is the basic application transformation and possible mappings on database level according to to the cardinality.  
\begin{table}
	\caption{The mapping between evolution of the application and database transformations. The  transformation connected with inheritance are in Sect.\ref{sec:inheritance}. }
	\label{tab:sw-basic-evolution}
\centering
	\begin{tabular}{ll}
		\hline
	Application Transformation & Database Transformation \\
	\hline
	\\[-2ex] 
	newProperty & \textbf{if} $cardinality \leq 1$ \textbf{then} addColumn \textbf{else} addTable \\

@@

\subsection{Advanced Evolutionary Transformations}
\label{sec:sw-adv-evolution}
Advanced evolutionary transformations are based on the basic ones, they can be obtained as a concatenation of transformations. This means the complex transformations are limited in the same way as their basic components. All transformations from the set $E$ presented so far has the same input information for both application and database, whereas the advanced transformations usually needs a mapping between stored data (instances) as theirs input.

\paragraph{Copy Property}
The $copyProperty$ creates a duplicate of a property in a given class. It can rename the property so it can be used for creating duplicates of a property in the property owning class. If the mapping is not provided the transformation creates only a structural copy of the property otherwise it copies the values too. Therefore the  $copyProperty$ transformation is the simplest way to manipulate stored data.
\begin{align}
& copyProperty(sourceClass, propertyName, targetClass, \nonumber  \\
& \; \; \; mapping, software) 
\end{align}
The  $copyProperty$ is the first transformation when an additional information has to be added to the usual code refactoring. It is because the $copyColumn$ transformation needs one more information to succeed - there has to be provided mapping between the source and target table to assure the data information consistency. %Moreover the source and target should have the same structure, type and constraints.
The transformation is interpreted for both software components - in case of the application a new property is added:
\begin{align}
& \Psi(copyProperty(sourceClass, property, targetClass, mapping, \nonumber  \\ 
&  \; \; \; software)) = newProperty(targetClass, property, application) 
\end{align}
in case of database the copy of a column or table is created according to property's cardinality and then values are copied:
\begin{align}
& \Phi(copyProperty(sourceClass, property, targetClass, newName, \nonumber \\ 
&  \; \; \; mapping, software)) = \begin{cases}
 coypColumn(ORM(sourceClass),  \\ \;\;\; ORM(targetClass),  \\ \;\;\;alterColumnName(ORM(property), \\ \;\;\; ORM(newName)), mapping, database) \\ \; \mathbf{if} \; cardinality(property) \leq 1 \\\\
 copyTable(ORM(Property), \\ \;\;\; ORM(newName))
 \end{cases}
\end{align}
The \emph{copyProperty} transformation gets a mapping between instances (stored data) on input. The mapping to provides information which value from the source table rows is copied to which row in target table. The mapping is usually defined as: 
\begin{align}\label{eq:mapping}
&	mapping : select(ORM(sourceName), database) \nonumber \\ 
& \;\;\;\;\; \rightarrow  (select(ORM(targetName), database) 
\end{align}
However different definitions can be used the only condition is the result of the mapping has to respect the structure of the instances and constraints on column (e.g. $NULL$ cannot be inserted into a column with $NOTNULL$ constraint defined) The character of transformation depends on the feature of provided mapping. There occurs no data lost if the mapping is surjective.

\paragraph{Move Property}
The $moveProperty$ transformation is based on the $copyProperty$ transformation followed by the $removeProperty$ so its construction is easy. On the other hand special attention has to be paid to the provided mapping of instances, because it can cause lost of data. The ideal case is when the mapping is surjective, then the transformation can not cause the lost of data. %On the other hand the situation when the mapping does not provides pairs for all rows in a source table can lead to data loss. 
\begin{align}
& moveProperty(sourceClass, property, targetClass, newName,\nonumber  \\
& \;\;\; mapping, software) = removeProperty(sourceClass, property, \nonumber \\
& \;\;\; copyProperty(sourceClass, property, targetClass, \nonumber \\ 
& \;\;\;newName, mapping, software))
\end{align}

>>>>

\title{Application and Relational Database Co-Refactoring}

@@

\begin{abstract}
A change of application source code caused by refactoring affects not only the application, but it affects the stored data too if the persistent objects change. The change is usually processed in two steps: refactoring and data migration. We provide a solution capable to migrate database according to refactoring in the application code. The feasibility of the change and its data-secure processing is addressed too.

@@

\subsection{Solution Proposal}
We believe that there exists a better solution which automatizes the process of database evolution according to the code refactoring. The solution is illustrated in the Fig.~\ref{fig:evolution}. It is based on a change in the evolution process which assumes that the ORM does not change during the evolution:
\begin{enumerate}
	\item The evolution of the whole software is defined independently on the application or the database.
	\item The evolution is interpreted for the application and the database.
	\item The evolution is executed.
\end{enumerate}
The process can decrease human mistakes in the evolution process, because there is only one source for the evolution and the evolution is automatically interpreted for the application (as a refactoring) and database (as schema and data migration). The existence of the set of all possible software evolutions $E$ is based on a evolutionary transformations specific for application and database. Each transformation contains conditions of transformation feasibility, thus the feasibility of the evolution can be verified.

@@

\paragraph{Example 2} The same situation from Example 1 can be solved more effectively when the \emph{inline} evolution is defined in $E$. All information needed for software evolution is provided as the input of the \emph{inline} transformation which is then automatically interpreted as inline of application classes and merge of tables on database level. The mapping is provided as the input of \emph{inline} too, thus the script for data migration can be generated. The structural feasibility is verified during the interpretation of the \emph{inline} and the migration script contains conditions which verifies the feasibility on the level of data.

@@

\section{Model of Software Evolution}
\label{sec:models}
Model-driven development (MDD) is a good approach to data evolution of various software components \cite{Moravec:PracticalApproach}. The software is represented by a set of models -- an application model and a database model concretely. The evolution and ORM is represented by a set of model-to-model transformations and the interpretation of transformations from $E$ is a model-to-model transformation again.

Three sets of transformations has to be defined to provide the capabilities described in Example 2: the set $E$ of all possible software transformations:
\begin{equation}
E = \{t | t: Software \rightarrow Software\}\:,
\end{equation}
the set of application refactorings:
\begin{equation}
A = \{t | t: Application \rightarrow Application\}
\end{equation}
and the set of database evolutionary transformations
\begin{equation}
 D = \{t | t: Database \rightarrow Database\}\:.
\end{equation}
Evolution of the software is defined as an interpretation of the transformation for each component of the software:
\begin{align}
& t(s) = \begin{cases}
software(\Psi(s, t), \Phi(s,t)) \; \mathbf{if} \; s \neq \perp \\\\
 	\perp
\end{cases}
s \in Software, t \in E \nonumber
\end{align}
Where $\Psi : E \rightarrow A $ interprets the software evolution cases to the code refactoring and $\Phi : E \rightarrow D $ interprets to the evolutionary transformation of database. The $\perp$ symbol denotes an inconsistent state of a software or its component. The definition of evolution allows to concatenate evolutionary transformations easily.

\subsection{Software Model}
The software consists of three important components (as seen in Fig. \ref{fig:evolution}: an application (its persistent layer concretely), a database (consisting of a database schema and stored data) and an ORM, therefore we define a software as a triple consisting of an application and a database which are connected together by an ORM.
The software has to be in consistent state so its users could benefit from its usage. The software is in a consistent state if the application and the database are consistent and the database structure corresponds to the application structure according to the ORM, therefore the software is defined as:
\begin{align}
& software(a, d, \rho) = \begin{cases}
 software(a, d, \rho) \; \textbf{if} \; a \neq \perp \; \wedge \; b \neq \perp \\ \wedge \; \rho(a) = schema(d) 
 \\\\
 \perp
 \end{cases}\\
& a \in Application, d \in Database, \rho \in ORM \nonumber
\end{align}
Evolution of a software is a transformation from one consistent state to another one. These states are called generations of the software and the function which changed the state of the software are called transformations.

@@

\subsection{Application Model}
An application is defined as a sequence of classes and it creates the context for all structures used in the software persistent layer:
\begin{align}
& \mathbf{Application} = (Class*) \\
& 	\mathbf{Class} = (label, Property*, Association*, Inheritance) \\
& \mathbf{Property} = (label, AppType, DefaultValue, \nonumber \\ & \;\;\;\;  Cardinality, Mandatory) \\
&	\mathbf{Association} = (label, classRef, startCardinality, endCardinality)  \\
& \mathbf{Inheritance} = (class, InheritanceType) \\¨
& \mathbf{InheritanceType} = SINGLETABLE \; | \; TABLE-PER-CLASS \\
& \mathbf{AppType} = APPSTRING
\end{align}

\paragraph{Class} Class represents a basic structural unit in the application model. It has a unique name, one or more properties and it can be associated to other classes in the application. The properties of a class can be of two kinds - values or collections - according to their cardinality.

\paragraph{Property} Property represents a feature of a class which is represented as a primitive type. A property can be mandatory, can have a default value and according to its cardinality it can represent a single value or a collection of values.

\paragraph{Association} Association represents a connection between two classes. It has a unique name and the reference is represented by the label of referenced class. The class which owns the association is consider to be the starting class of an association, referenced class is consider to be the ending class of an association. The cardinalities defines the multiplicity of both association ends.

\paragraph{Inheritance} 
%TODO Inheritance
Multiple ancestors are not allowed and of course a class can not be its own ancestor cyclically. We also assume there is only one type of inheritance type in the class hierarchy.

@@

\subsection{Application Manipulation}
\label{sec:app-evolution}
The application model defines only the structure of an application thus there are defined transformations for its creating, altering and removing parts of the model (the set $A$). The conditions for successful transformations are simple in case of creating or altering:
\begin{enumerate}
	\item The name collisions have to be prevented when creating a class in the model or altering its name.
	\item All references have to be updated when renaming a class.
	\item The existence of referenced class has to be verified when creating or altering an association.  
\end{enumerate}
The only problem during deleting classes etc. from an application is  that the class can be removed only in case it is not associated by other classes. The transformations from $A$ are in Table \ref{tab:app-evolution}. The list of altering operations is not complete as transformation for changing the obligation or cardinality of properties and associations are missing. It is because these transformations are not so important for complex refactorings. 
\begin{table}
\caption{The transformations for application evolution.}
	\label{tab:app-evolution}
\centering
	\begin{tabular}{ll}
	\hline
	Type of transformation & Defined transformations \\
	\hline
	\\[-2ex] 
	Application Creation
	& newProperty(className, property, application) \\
	& newAssociation(className, association, mapping, \\ &  \hspace{0.5in}application) \\
	&  newClass(class, application) \\
	Application Alternation
		& renameProperty(className, oldName, newName, \\ &  \hspace{0.5in} application) \\
		& renameAssociation(className, oldName, newName, \\ &  \hspace{0.5in}application) \\
		& renameClass(oldName, newName, application)\\
		& changeReference(referncedClass, newReferencedClass) \\
		Application Disassembly 
		& removeProperty(className, propertyName,\\ &  \hspace{0.5in} application) \\
		& removeAssociation(className, associationName, \\ &  \hspace{0.5in}application) \\
		& removeClass(className, application)\\
%	Copy Structure
%		& copyProperty(sourceClass, property, targetClass,\\ &  \hspace{0.5in} newName, application)  \\
%		& copyClass(class, newName, application) \\
	Inheritance manipulation   & addParent(class, parentClass) \\
		& removeParent(childClass) \\
		& pushDown(class, property)\\
		& pullUp(class, property)\\
	\hline
	\end{tabular}
\end{table}

\subsection{Database Model}
A relational database consists of a database schema which defines the structure of the database and data, which represents stored instances in the software. A database is defined as:
\begin{align}
&	\mathbf{Database} = ( Table*, Row* )\\
&	\mathbf{Table} = (label, primaryKey, Column*, ForeignKey*)\\
&	\mathbf{Column} = (label, DbType, DefaultValue, Constraint*) \\
&	\mathbf{ForeignKey} = (Label, Label, Constraint*) \\
&	\mathbf{PrimaryKey} =  ( Label, Sequence ) \\
&	\mathbf{DbType} = DBSTRING \; | \; DBINT\\
&	\mathbf{Constraint} = NOTNULL \; | \; UNIQUE 
\end{align}

\paragraph{Table} Table represents a basic concept of a database schema. It has a unique name, one or more columns and it can be related to other tables in the schema by foreign keys. Rows in the table represents stored data.

@@

\paragraph{Primary key} Primary key is unambiguous identifier of a record in a table. The primary key is always provided (automatically generated) by the database as a non-zero natural number. Primary key is always defined with constraints\\ $NOTNULL$ and $UNIQUE$. The sequence generates values of primary key. A new value of a key is obtained by calling the function $next(s)$. The sequence which provides numbers incremented by one is called $s_1$ when the increment is 1, $s_{10}$ when the increment is 10 etc.

@@

\paragraph{Data} A database consists not only of a schema but also of data which are represented as rows in a table. A table row in our model is a tuple consisting of value pairs, which represent concrete value of a concrete column or key. Each row contains a name of a table it belongs to.
\begin{align}
&	\mathbf{Row} = (Label, Pair*) \\
&	\mathbf{Pair} = (Label, Value) 
\end{align}

\subsection{Database Manipulation}
\label{sec:db-evolution}
A database consists of two parts - of a database schema, which defines the structure, and of stored data - hence the transformations from set $D$ have to consider both parts. The transformation for manipulation of structure has similar conditions to the evolution of application. The transformations for data manipulation are inspired by the SQL language. The basic transformations are in Table \ref{tab:db-basic-evolution}. The set of transformations for database manipulation is limited in contrast to the SQL language. Transformations only necessary for data evolution on database level are introduced.

\begin{table}
\caption{The transformations for database evolution. Transformations can be mapped to SQL intuitively.}
	\label{tab:db-basic-evolution}
\centering
\setlength\tabcolsep{0.5em}
	\begin{tabular}{ll}
	\hline
	Type of transformation & Defined transformations \\
	\hline
	\\[-2ex] Database Creation
	& addColumn(tableName, column, database) \\
	& addForeignKey(tableName, key, database) \\
	& addTable(table, database)\\
	Database Alternation
	& alterColumnName(tableName, oldName, \\ & \hspace{0.5in}newName, dataabse) \\
	& alterForeignKeyName(tableName, oldName, \\ & \hspace{0.5in}newName, dataabse) \\
	& alterTableName(oldName, newName, dataabse) \\
	Database Disasembling
	& dropColumn(tableName, columnName, database) \\
	& dropForeignKey(tableName, keyName, database) \\
	& dropTable(tableName, database) \\
	Data Manipulation
	& select(tableName, rowId, database) \\
	& select(tableName, database) \\
	& insert(row, database) \\
	& insert(rowId, value, database) \\
	& delete(tableName, rowId, database) \\
	& delete(tableName, rowId, columnName, database) \\
	Copy Structure
	& copyColumnStructure(sourceTableName,\\ &  \hspace{0.5in} targetTableName,  sourceColumnName, \\ &  \hspace{0.5in} newName, database) \\
	& copyTableStructure(sourceTable, newTableName, \\ &  \hspace{0.5in} database) \\
	Copy Structure and Values
	& copyColumn(sourceTable, targetTable, \\
	& \hspace{0.5in} sourceColumnName, targetColumnName, \\ 
	& \hspace{0.5in} mapping, database) \\
	& copyTable(sourceTable, newName, database) \\
	Data-secure Database Elements &
	dropEmptyColumn(tableName, columnName, \\
	Removal & \hspace{0.25in} database) \\
	& dropEmptyForeignKey(tableName, keyName, \\ & \hspace{0.25in} database) \\
	& dropEmptyTable(tableName, database) \\
	\hline
	\end{tabular}
\end{table}

\paragraph{Copying Database Elements} The transformations for copying the structure and values of a column or table serve more as helpers for advanced evolution cases, where they are discussed in detail.

@@

\paragraph{Data-secure Database Elements Removal} The transformation for remove elements from the database can have fatal impact on the data preservation, therefore the set of transformations is extended by data secure transformations for removing database elements. These transformations are not a part of the SQL standard, although they can be implemented as database functions. These transformations create a secure way to remove elements from the database as they drop empty structural elements only.

\subsection{Object Relational Mapping}
ORM is the only fixed point in the software model we use. The mapping is similar to the Hibernate mapping \cite{Hibernate}, thus a lot of developers should be familiar with it. The main ideas are: 
\begin{itemize}
	\item classes are mapped to tables,
	\item single properties are mapped to a column or if the property is a collection then the property is mapped to a table,
	\item associations are mapped to foreign keys or to tables if the association represents many-to-many  relationship,
	\item a primary keys are created automatically for each table,
	\item names used in application are mapped into the database schema (e.g. because of possible name collision of application classes and names of database schema elements) 
\end{itemize}
%The ORM mapping is a function $ORM : Application \rightarrow Database $.

@@

\subsection{Basic Evolutionary Transformations}
\label{sec:sw-basic-evolution}
The evolution of the whole software is based on the atomic transformations specific for each software part, which are defined in Sec. \ref{sec:app-evolution} and \ref{sec:db-evolution}. This section introduces how those primitives can be used to manipulate the software elements.

The basic evolutionary transformations of the software are based on basic evolution of an application. These transformations has to respect the ORM, because properties and associations can be mapped as a column (foreign key respectively) as shown in the example of creating a new property: 
\begin{align}
& \Phi(newProperty(className, property, application)) = \nonumber \\
& \;\;\; =  \begin{cases}
  addColumn(ORM(className), ORM(property), database) \\\mathbf{if} \; cardinality(property \leq 1)  \\\\
  addTable(ORM(property), database) \\
  \mathbf{if} \; cardinality(property > 1)  
   \end{cases}
\end{align}

Mapping between basic evolutionary transformations is provided in the shortened way in Table \ref{tab:sw-basic-evolution}. There is the basic application transformation and possible mappings on database level according to the cardinality.  
\begin{table}
	\caption{The mapping between evolution of the application and database transformations. The  transformation connected with inheritance are in Sect.\ref{sec:inheritance}. }
	\label{tab:sw-basic-evolution}
\centering
	\begin{tabular}{ll}
		\hline
	Application Transformation & Database Transformation \\
	\hline
	\\[-2ex] 
	newProperty & \textbf{if} $cardinality \leq 1$ \textbf{then} addColumn \textbf{else} addTable \\

@@

\subsection{Advanced Evolutionary Transformations}
\label{sec:sw-adv-evolution}
Advanced evolutionary transformations are based on the basic ones, they can be obtained as a concatenation of transformations. This means the complex transformations are limited in the same way as their basic components. All transformations from the set $E$ presented so far have the same input information for both an application and a database, whereas the advanced transformations usually need a mapping between stored data (instances) as their input.

\paragraph{Copy Property}
The $copyProperty$ creates a duplicate of a property in a given class. It can rename the property so it can be used for creating of duplicates of a property in the property owning class. If the mapping is not provided the transformation creates only a structural copy of the property, otherwise it copies the values too. Therefore the  $copyProperty$ transformation is the simplest way to manipulate the stored data.
\begin{align}
& copyProperty(sourceClass, propertyName, targetClass, \nonumber  \\
& \; \; \; mapping, software) 
\end{align}
The  $copyProperty$ is the first transformation when an additional information has to be added to the usual code refactoring. It is because the $copyColumn$ transformation needs one more information to succeed - there has to be provided mapping between the source and the target table to assure the data information consistency. %Moreover the source and target should have the same structure, type and constraints.
The transformation is interpreted for both software components - in case of the application a new property is added:
\begin{align}
& \Psi(copyProperty(sourceClass, property, targetClass, mapping, \nonumber  \\ 
&  \; \; \; software)) = newProperty(targetClass, property, application) 
\end{align}
In case of database the copy of a column or table is created according to property's cardinality and then values are copied:
\begin{align}
& \Phi(copyProperty(sourceClass, property, targetClass, newName, \nonumber \\ 
&  \; \; \; mapping, software)) = \begin{cases}
 coypColumn(ORM(sourceClass),  \\ \;\;\; ORM(targetClass),  \\ \;\;\;alterColumnName(ORM(property), \\ \;\;\; ORM(newName)), mapping, database) \\ \; \mathbf{if} \; cardinality(property) \leq 1 \\\\
 copyTable(ORM(Property), \\ \;\;\; ORM(newName))
 \end{cases}
\end{align}
The \emph{copyProperty} transformation gets a mapping between instances (stored data) on input. The mapping provides information which value from the source table rows is copied to which row in the target table. The mapping is usually defined as: 
\begin{align}\label{eq:mapping}
&	mapping : select(ORM(sourceName), database) \nonumber \\ 
& \;\;\;\;\; \rightarrow  (select(ORM(targetName), database) 
\end{align}
However, different definitions can be used, the only condition is the result of the mapping has to respect the structure of the instances and constraints on column (e.g. $NULL$ cannot be inserted into a column with $NOTNULL$ constraint defined). The character of transformation depends on the feature of provided mapping. There occurs no data lost if the mapping is surjective.

\paragraph{Move Property}
The $moveProperty$ transformation is based on the $copyProperty$ transformation followed by the $removeProperty$ so its construction is easy. On the other hand, special attention has to be paid to the provided mapping of instances, because it can cause lost of data. The ideal case is when the mapping is surjective, then the transformation can not cause loss of data. %On the other hand the situation when the mapping does not provides pairs for all rows in a source table can lead to data loss. 
\begin{align}
& moveProperty(sourceClass, property, targetClass, newName,\nonumber  \\
& \;\;\; mapping, software) = removeProperty(sourceClass, property, \nonumber \\
& \;\;\; copyProperty(sourceClass, property, targetClass, \nonumber \\ 
& \;\;\;newName, mapping, software))
\end{align}

\paragraph{Extract and Inline Class}
Extract and inline class are two opposite transformations. First of them extracts a new class from an existing class, second of them moves data from one class into another. Both transformations can be composed from already mentioned basic transformations:
\begin{align}
& extract(sourceClass, newClass, property*, software) = \nonumber  \\ 
& \; \; \; = moveProperty(sourceClass, newClass, property*, mapping \nonumber \\ 
& \; \; \; addClass(newClass, software))
\end{align}
The \emph{mapping} cannot be created according to the Definition \ref{eq:mapping}, because there are no stored data in the table representing the \emph{newClass}, therefore the \emph{mapping} is defined as identity in case of \emph{extract}, therefore the rows in the new table has same primaryKey value as in their source class. The \emph{extract} transformations assumes the \emph{sourceClass} and \emph{newClass} have the same type of primary keys sequence. 
\begin{align}
& inline(targetClass, classToInline, mapping, software)) =  \nonumber \\
& \; \; \; = removeClass(classToInline, moveProperty(classToInline, \nonumber \\
& \; \; \; targetClass, properties(classToInline), mapping, software))
\end{align}
